---
title: "Mini-Project 02"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

## Mini Project 2 Background

The second Mini Project focused on real data collected via wearable devices from a fellow FPU Graduate student. While the project he is focusing on uses machine learning techniques to predict risk of hypertension in young adults - I sought to provide meaningful visualizations from the data collected via the wearable devices or participants themselves.


```{r setup, include=FALSE}

library(ggplot2)
library(tidyverse)
library(dplyr)
library(lubridate)
library(scales)
library(viridis)
library(ggthemes)
library(sf)

surv_join_zip <- read.csv("https://raw.githubusercontent.com/jordanrosedouglass/hypertension_research/master/survey_data_w_zip.csv")

master_collection <- read.csv("https://raw.githubusercontent.com/jordanrosedouglass/hypertension_research/master/JuneJulyAugMaster2020.csv")

mergesurvey <- merge(master_collection, surv_join_zip, by=c('MRN'), all.x=T)

# ZIPCODE needs to be read as character here
surv_join_zip$ZIPCODE <- as.character(surv_join_zip$ZIPCODE)

#read the shapefile into .rmd
fl_shapes <- read_sf("/Users/jordandouglass/Downloads/dataviz_mini-project_02/data/kx-florida-zip-code-areas-SHP/florida-zip-code-areas.shp")

#count the frequency of ZIPCODES appearing in the survey data
mydf <- count(surv_join_zip,ZIPCODE)

participant_map <- fl_shapes %>%
  left_join(mydf, by = "ZIPCODE")

```

# Datasets

This project required three datasets - one of which is an .`shp` file containing Florida Zip Code geometries from __[Koordinates](https://koordinates.com/)__ two of which are pulled from __[my very unorganized GitHub Repo](https://github.com/jordanrosedouglass/hypertension_research)__. After using several iterations of the `join()` function (none of which appeared to be of my liking), the `merge()` function was used to _back-door_ join the datasets.

Through a great deal of frustration and stress, eventually, I was able to use the `join` function for the geometry data to join the Zip Codes of participants. :sweat_smile:

## Visualizing Spatial Features

When first creating the map, I very quickly realized just how  small zip codes were - they are __really small in Florida__. This lead to some issues in terms of visualization - the map of Florida was great - the filling of participants by their zip codes? _Not so much.._ This was eventually resolved by deciding to use an interactive map - we can zoom in to see the variation of participants in the Central Florida area.

```{r mapplot}
library(ggrepel)
#create map filling by frequency of ZIPCODE appearing
pmap <- participant_map %>% 
  ggplot()+
  geom_sf(aes(fill = n), color = "white") +
  theme_light() +
    guides(fill= guide_legend(reverse=TRUE)) +
  theme(aspect.ratio = 0.8, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  labs(title="BioDatix Participants", caption = "Source: BioDatix Participant Data") +
  labs(fill="Participants") +
  scale_fill_continuous()

library(plotly)
library(htmlwidgets)

p <- ggplot(participant_map) + geom_sf(aes(fill=n))

ggplotly(p) %>%
  highlight(
    "plotly_hover",
    selected = attrs_selected(line = list(color = "white"))
  )
```

## Plotting Variables

Following the map plots - we wanted to visualize the __systolic__ and __diastolic__ blood pressure and __heart rate__ from the participants; this will help to validate the variables collecting the data the way they are supposed to.


For the purpose of organization, a new dataframe `long` was created to convert the wide data to long format - this way every row represents an observation from a category; we have also removed missing values from `BP_Sys`, `BP_Dia`, and `HeartRate`. We have also created variables indicating what is considered _high_ blood pressures and heart rates.

From the plot, there is a linear relationship: as heart rate increases, so does blood pressure - this is to be expected as heart rate and blood pressure are intimately related.



```{r plot_bp}

master_collection <- mergesurvey %>% 
  filter(!is.na(BP_Sys), !is.na(BP_Dia), !is.na(HeartRate)) %>% 
  mutate(
    blood_pressure = ifelse(BP_Sys > 130 | BP_Dia > 90, "High", "Normal"),
  h_rate = ifelse(HeartRate >= 145, "High", "Normal"),
  Gender = as.factor(Gender))

long = master_collection %>%
  select(BP_Dia, BP_Sys, HeartRate,Gender ) %>% 
  gather(bp, value, BP_Dia:BP_Sys) %>% 
  mutate(bp = recode(bp, `BP_Dia` = "Diastolic", `BP_Sys` = "Systolic"))



bpplot <- ggplot(long, aes(value, HeartRate, color = bp)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method= "lm",
              formula = "y ~ x",
              colour="#479FD0") +
facet_grid(~bp) + 
  scale_color_discrete() +
  theme_minimal() +
  theme(aspect.ratio = 0.8, panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(text = element_text(family = "serif", size = 11), legend.position="none") +
  xlab("Blood pressure (mm Hg)") +
  ylab("Heart Rate") +
  ggtitle("Blood pressure vs. Heart Rate") +
  labs(
    caption = "Source: Biodatix Wearable Data")

bpplot

```


## Gender and Blood Pressure

The following visualization was further faceted by `gender`. Here, we can clearly see that there is not much difference between gender and blood pressure - however noticed is there are more instances of outliers in female than male participants.

```{r linear model formula}
ggplot(long, aes(value, HeartRate, color = Gender)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method= "lm",
              formula = "y ~ x",
              colour="#479FD0") +
  facet_grid(bp ~ Gender) +
  scale_color_discrete() +
  theme_hc() +
    theme(text = element_text(family = "serif", size = 11), legend.position="none") +
  xlab("Blood pressure (mm Hg)") +
  ylab("Heart Rate (BPM)") +
  ggtitle("Blood pressure vs. Heart Rate vs. Gender") +
  labs(caption = "Source: Biodatix Wearable Data")

```








